Process-based phenotypes: an application to the Cellular Phenotype Ontology

### 2011-11-22 paper/drinking.tex

A first kind of abnormality of regulatory processes are {\em abnormal
  temporal distribution patterns} of a process. The most common
abnormal distribution pattern

We use PATO's {\em frequency} class to define {\em Abnormal frequency
  of occurrence of $X$} as a subclass of an {\em Abnormality of the
  regulation of $X$}.

>>>>

A first kind of abnormality of regulatory processes are {\em abnormal
  temporal distribution patterns} of a process. In these
abnormalities, the {\em way} in which processes of a particular kind
are temporally distributed is abnormal.  The most common abnormal
distribution pattern is an increased or decreased frequency, and we
use PATO's {\em frequency} class to define {\em Abnormal frequency of
  occurrence of $X$}.
% \begin{verbatim}
% has-phenotype some (has-part-some (participates-in some 
%   (regulates some X and has-quality some (frequency and towards some X))))
% \end{verbatim}
For example, an {\em Abnormal frequency of occurrence of apoptosis} is
defined as an abnormality of {\em Regulation of apoptosis} ({\tt
  GO:0042981}) with respect to the {\em frequency} ({\tt
  PATO:0000044}) of {\em Apoptosis} ({\tt GO:0006915}) occurrences.

There are further types of deviation from a distribution pattern. For
example, a kind of process that is normally {\em rhythmic} can be
abnormal in that it is {\em arrhythmic}. A typical example of this
kind of process is {\em Heart beating} ({\tt GO:0060047}), in which
{\em Cardiac muscle contraction} ({\tt GO:0060048}) processes occur in
a rhythmic pattern. In {\em Cardiac dysrhythmia}, however, {\em
  Cardiac muscle contraction} processes occur arrhythmically, and we
consider this to be an abnormality of the regulation of {\em Cardiac
  muscle contraction}. While these abnormalities are often highly
informative in clinical diagnostics and biological investigations, we
usually lack the necessary information that is required to
automatically determine meaningful types of abnormal distribution
patterns.

A second kind of regulatory abnormalities is related to the {\em
  onset} of a process. With respect to a reference process, a
particular kind of process may be {\em delayed} ({\tt PATO:0000502})
or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
  apoptosis} refers to an abnormality of the {\em Regulation of
  apoptosis} in which apoptosis is induced later than normal.
We use
